---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
  
---
##As requested we need to set the default of echo to be true.
```{r}
library(knitr)
opts_chunk$set(echo = TRUE)
```

##Loading required libraries
```{r}
library(dplyr)
library(lubridate)
library(ggplot2)
```

## Loading and preprocessing the data (Point 1 of assignment)

```{r}
if(!file.exists('activity.csv')){
    unzip('activity.zip')
}
activityData <- read.csv('activity.csv')
```
## Process/transform the data (if necessary) into a format suitable for your analysis
```{r}
data <- read.csv("activity.csv", header = TRUE, sep = ',', colClasses = c("numeric", "character", "integer"))

data$date <- ymd(data$date)
```
## We can check the data with str() and head() for confirmation.
```{r}
str(data)
head(data)

```
## What is mean total number of steps taken per day?
## We use dplyr to calculate total number of steps per day and group by dates.
```{r}
steps <- data %>%
  filter(!is.na(steps)) %>%
  group_by(date) %>%
  summarize(steps = sum(steps)) %>%
  print
```

## Create histogram (Point 2 of assignment)
```{r}
ggplot(steps, aes(x = steps)) +
  geom_histogram(fill = "firebrick", binwidth = 1000) +
  labs(title = "Histogram of Steps per day", x = "Steps per day", y = "Frequency")
``` 

## Calculating Mean and Median (Point 3 of assignment)
```{r}
mean_steps <- mean(steps$steps, na.rm = TRUE)

median_steps <- median(steps$steps, na.rm = TRUE)

median_steps
## [1] 10765
mean_steps
## [1] 10766.19


```

## What is the average daily activity pattern?

## Lets first calculated the average number of steps taken in each 5 minute interval using dplyr and group by interbal
```{r}
interval <- data %>%
  filter(!is.na(steps)) %>%
  group_by(interval) %>%
  summarize(steps = mean(steps))
```  
## Now we can use ggplot for the 5 minute interval and average steps taken time series. (Point 4 of Assignment)
```{r}
  ggplot(interval, aes(x=interval, y=steps)) +
  geom_line(color = "firebrick")
```
  
## The 5-minute interval that, on average, contains the maximum number of steps (Point 5 of assignment)
```{r}
  interval[which.max(interval$steps),]

## Imputing missing values (Point 6 of assignment)

## First lets sum all missing values

sum(is.na(data$steps))
## [1] 2304
## so we have 2304 missing values
## we will fill these missing values with the average number of steps in the same 5 minute interval.

data_full <- data
nas <- is.na(data_full$steps)
avg_interval <- tapply(data_full$steps, data_full$interval, mean, na.rm=TRUE, simplify=TRUE)
data_full$steps[nas] <- avg_interval[as.character(data_full$interval[nas])]

##Check that there are no missing values:
## we can confirm that there are no missing values in the new dataset we created.

sum(is.na(data_full$steps))

## [1] 0
## so there are no missing values now.
```

## Now Lets Calculate the number of steps taken in each 5-minute interval per day using dplyr and group by interval. Use ggplot for making the histogram:

```{r}
steps_full <- data_full %>%
  filter(!is.na(steps)) %>%
  group_by(date) %>%
  summarize(steps = sum(steps)) %>%
  print
 
##Histogram of the total number of steps taken each day after missing values are imputed (Point 7 of assignment)
  ggplot(steps_full, aes(x = steps)) +
  geom_histogram(fill = "firebrick", binwidth = 1000) +
  labs(title = "Histogram of Steps per day, including missing values", x = "Steps per day", y = "Frequency")

mean_steps_full <- mean(steps_full$steps, na.rm = TRUE)

mean_steps_full
## [1] 10766.19

median_steps_full <- median(steps_full$steps, na.rm = TRUE)
median_steps_full
## [1] 10766.19
```

## Are there differences in activity patterns between weekdays and weekends?
```{r}
data_full <- mutate(data_full, weektype = ifelse(weekdays(data_full$date) == "Saturday" | weekdays(data_full$date) == "Sunday", "weekend", "weekday"))
data_full$weektype <- as.factor(data_full$weektype)
head(data_full)

## Lets Calculate the average steps in the 5-minute interval and use ggplot for making the time series of the 5-minute interval for weekday and weekend, and compare the average steps:

interval_full <- data_full %>%
  group_by(interval, weektype) %>%
  summarise(steps = mean(steps))
  
s <- ggplot(interval_full, aes(x=interval, y=steps, color = weektype)) +
  geom_line() +
  facet_wrap(~weektype, ncol = 1, nrow=2)
print(s)
```